### 2015-01-14 doc/thesis/system.tex
Moving the discussion about multiple assignments before the discussion about refinement of table types

Multiple assignments also appear in function and method applications,
but our type system needs different typing rules for
expression lists that appear in function and method applications,
as we also want to catch arity mismatch.
For instance, assuming that $x$ and $f$ are in the environment with
types $\Integer \cup \Nil$ and
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void \rightarrow \Integer{*}$,
respectively.
We want that
\[
x = f(``foo")_{m}
\]
type checks because
\[
``foo" \times \Nil \times \Nil \times \Void \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]
but we also want that
\[
x = f(``foo",1,2,3)_{m}
\]
does not type check because
\[
``foo" \times 1 \times 2 \times 3 \times \Void \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]

To do that, the rule \textsc{T-APPLY1} uses rules that type check
expression lists, but that do not include $\Nil{*}$ in the end of
the tuple type.
Instead, it uses the predicate $adjust$ to replace any missing
value with $\Nil$:
\[
\begin{array}{c}
\mylabel{T-APPLY1}\\
\dfrac{\env_{1} \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
       \env_{2} \vdash el:r_{3}, \env_{3} \;\;\;
       r_{4} = adjust(r_{3}, r_{1}) \;\;\;
       r_{4} \lesssim r_{1}}
      {\env_{1} \vdash e(el)_{m}:r_{2}, \env_{3}}
\end{array}
\]

@@

We start with the rule that defines the most unusual feature of
our type system: the refinment of table types.
The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
and \emph{unique} table types:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\begin{array}{c}
       \env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique}\\
       \env_{1} \vdash c:t_{1}, \env_{2} \;\;\;
       \not \exists i \in 1..n \; t_{1} \lesssim k_{i} \;\;\;
       \Nil \not\lesssim t_{1} \;\;\;
       t_{2} = close(t)
       \end{array}}
      {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t_{2}\}_{open|unique}]}
\end{array}
\]

@@

This example type checks because all the environment changes are consistent, and
$``foo" \times 1 \times \Nil{*} \lesssim \String \times \Integer \times \Value{*}$.
The type of the expression list includes the type $\Nil{*}$ in the end
because our type system has typing rules to add this type,
when the type of the expression list does not end with a variadic type.
Our type system has this behavior to use $\Nil$ in the place of missing
values, like Lua does.
This example uses the rule \textsc{T-EXPLIST2}, which is really close
to the rule \textsc{T-LHSLIST}:
\[
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\env \vdash e_{k}:t_{k}, \env_{k} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{e}\;|}
      {\env \vdash \vec{e}:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{m}}
\end{array}
\]

@@

Typed Lua also keeps track whether it is safe or not safe to
change the type of a table, but before discussing this subject,
we will discuss the different table constructors that our
abstract syntax includes, as they will appear in the next
examples.

Even though \emph{unique} table types overcome our limitation.
The covariance on mutable fields could allow programs that create unsafe
aliases to pass the type checker.
For instance,
\begin{verbatim}
    local t1 = { foo = 5 }
    local t2:{"foo":integer?} = t1
    t2.foo = nil
\end{verbatim}
does not type check, as the last line is erasing the field \texttt{foo}
from \texttt{t1}.
This program would type checks as it is, because our implementation uses
local type inference to assign static types to unannotated locals.
In this example, the type of \texttt{t1} would be subtype of the type of \texttt{t2}.
More precisely,
\[
\{``foo":\Integer\}_{unique} \subtype \{``foo":\Integer \cup \Nil\}_{closed}
\]

In the next section we will show in more detail that this example
does not type check, because our type system promotes an alias
expression from \emph{unique} to \emph{closed}, and promotes the
aliased variable from \emph{unique} to \emph{open}.
This means that the previous example does not type check because
the type of the expression \texttt{t1} is not subtype of the type
of the variable \texttt{t2}.
More precisely,
\[
\{``foo":\Integer\}_{closed} \not\subtype \{``foo":\Integer \cup \Nil\}_{closed}
\]

Thus, we introduced \emph{open} table types to handle the aliasing of table types.

The simplest form of the table constructor is the empty table $\{\}$.
Its inference rule is straightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env \vdash \{\}:\{\}_{unique}, \env
\end{array}
\]

Our abstract syntax reduces the more complex uses of the table
constructor into three forms: $\{\;\vec{p}\;\}$, $\{\;{...}_{m}\;\}$,
and $\{\;\vec{p},{...}_{m}\;\}$.
The first one uses a list of expression pairs $[e_{1}] = e_{2}$,
the second one uses just the vararg expression, and the third one uses both.
We did not include in the abstract syntax a table constructor
that uses a list of expressions because we can write it using the first form.
For instance, we can write the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.
We chose to not include this form because its inference rule is similar
to the inference rule of the first form, and just controls the type of
the omitted indices.
For the same reason we did not include table constructors with
multiple expressions, which can return a tuple type that does not end
with a variadic type.
We did include the vararg expression in the last position to show how
our type system can infer a table type that has a record part and
also an array part.

The inference rule \textsc{T-CONSTRUCTOR2} infers a table type for
the first form:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       t = wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique})}
      {\env_{1} \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
\end{array}
\]

This inference rule infers the type of each pair, uses these types
to build a table type, and uses the predicate \emph{wf} to check
whether the table type is well formed.
Formally, a table type is well-formed if it obeys the following rule:
\[
\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
\]

The inference rules for expression pairs would be straightforward
if our table types allowed any type in the keys.
The rules \textsc{T-PAIR1}, \textsc{T-PAIR2}, and \textsc{T-PAIR3}
check if the type of the key is a subtype of $\Boolean$, $\Number$,
or $\String$, respectively:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-PAIR1}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Boolean}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-PAIR2}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Number}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-PAIR3}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \String}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
\end{array}
\end{array}
\]

The rule \textsc{T-PAIR4} is a fallback for the previous rules,
but it works only if the type of the key does not include $\Nil$:
\[
\begin{array}{c}
\mylabel{T-PAIR4}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       \Nil \not\subtype t_{1}}
      {\env_{1} \vdash [e_{1}] = e_{2}: \Any{:}close(t_{2}), \env_{3}}
\end{array}
\]

The rule \textsc{T-CONSTRUCTOR2} along with the rules for expression
pairs allow Typed Lua to type check the following example:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\String:\Integer \cup \Nil\}_{closed} = \{ [``x"] = 1, [``y"] = 2 \} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; b:\Integer \cup \Nil = a[``x"] \; \mathbf{in} \; b = a[``z"]$}
\end{tabular}
\end{center}

This example type checks because Typed Lua infers the table type
$\{``x":1, ``y":2\}_{unique}$, which is subtype of
$\{\String:\Integer \cup \Nil\}_{closed}$.
Even though the key $``z"$ does not exist in the table $a$,
the two table indexations type check with type $\Integer \cup \Nil$,
which is subtype of the type of $b$.
The rule \textsc{T-INDEX1} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-INDEX1}\\
\dfrac{\env_{1} \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t, \env_{3} \;\;\;
       \exists i \in 1{..}n \; t \lesssim k_{i}}
      {\env_{1} \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
\end{array}
\]

Typed Lua handles arrays as hashes that maps integers to some type $t$.
In Lua, programmers often use the vararg expression to initialize arrays.
The rules \textsc{T-CONSTRUCTOR3} and \textsc{T-CONSTRUCTOR4} define
the behavior of the vararg expression for this case:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-CONSTRUCTOR3}\\
\dfrac{\env_{1}({...}) = t}
      {\env_{1} \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env_{1}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-CONSTRUCTOR4}\\
\dfrac{\begin{array}{c}
       \env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       \env_{i+1}({...}) = t_{v}\\
       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t_{v} \cup \Nil\}_{unique}) = t_{t}
       \end{array}}
      {\env_{1} \vdash \{\;\vec{p},\;{...}_{m}\;\}:t_{t}, \env_{n+1}}
\end{array}
\end{array}
\]

The rule \textsc{T-CONSTRUCTOR3} type checks the case where we use
only the vararg expression inside a table constructor to initialize
an array.
If we assume that the $...$ is in the environment and has type $\String$,
the following example type checks through this rule:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\Integer:\String \cup \Nil\}_{closed} = \{ {...}_{m} \} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; b:\String \cup \Nil = t[1] \; \mathbf{in} \; b = t[5]$}
\end{tabular}
\end{center}

The rule \textsc{T-CONSTRUCTOR4} type checks the case where we use
the table constructor as a record that includes an array part.
If we assume that the $...$ is in the environment and has type $\String$,
the following example type checks through this rule:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\Integer, ``y":\Boolean, \Integer:\String \cup \Nil\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``y"] = \mathbf{true}, {...}_{m} \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; a[``x"] = ``bar"$}
\end{tabular}
\end{center}
but the next example does not type check because the record part
includes fields that the key types are subtype of the array part:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{1:\String, 10:\Boolean, \Integer:\String \cup \Nil\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [1] = ``foo", [10] = \mathbf{true}, {...}_{m} \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; a[``x"] = ``bar"$}
\end{tabular}
\end{center}

Making the inference rules of the table constructor infer an \emph{unique}
table type also allow us to use them for initializing record types that
have optional fields.
We can include the optional field in the initialization
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\String, ``y":\String \cup \Nil, ``z":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``y"] = ``bar", [``z"] = ``baz" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}
and we can also omit it in the initialization
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\String, ``y":\String \cup \Nil, ``z":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``z"] = ``baz" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}

After discussing the inference rules of the table constructor,
we can get back to the discussion about the refinement of table types.

Besides allowing the refinement of table types through field assignment,
Typed Lua also keeps track whether it is safe or not safe to change the
type of a table.
%To do that, our type system always promotes an \emph{open}
%table to \emph{closed} and an \emph{unique} table to \emph{open}
%when they appear as expressions, or when they enter in a scope that is
%not its declaration scope.
%The order matters, as going from \emph{unique} to \emph{open}
%before going from \emph{open} to \emph{closed} closes \emph{unique} table types.
In the last section we mentioned that aliasing \emph{unique} table
types is unsafe.
To avoid unsafe aliasing of \emph{unique} table types, the rule \textsc{T-IDREAD1}
always promotes them to \emph{closed}, and updates the variable in the
environment to \emph{open}:
\[
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}}
      {\env_{1} \vdash n_{e}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}, \env_{1}[n \mapsto \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}]}
\end{array}
\]

This is the rule that makes the unsafe example from last section to
not type check:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{``foo":\Integer\}_{unique} = \{ [``foo"] = 5 \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``foo":\Integer \cup \Nil \}_{closed} = a \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$b[``foo"] = \mathbf{nil}$}
\end{tabular}
\end{center}

This examples does not type check because aliasing $a$ produces the
type $\{``foo":\Integer\}_{closed}$ that is not subtype of
$\{``foo":\Integer \cup \Nil\}_{closed}$, the type of $b$.

However, sometimes we want to use \emph{unique} table types to initialize
\emph{closed} table types, as they allow us to initialize optional fields.
To do that, Typed Lua includes a cast expression that allow us
to use an \emph{unique} table type before it becomes \emph{open}.
The rule \textsc{T-CAST} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-CAST}\\
\dfrac{\env_{1}(n) \subtype t}
      {\env_{1} \vdash {<}t{>} \; n:t, \env_{1}[n \mapsto t]}
\end{array}
\]

@@

We can use $a$ to initialize $b$ because the cast expression converts
the type of $a$ from $\{``x":\String, ``y":\String\}_{unique}$ to
$\{``x":\String, ``y":\String \cup \Nil\}_{open}$, which is subtype of
$\{``x":\String, ``y":\String \cup \Nil\}_{closed}$.
We can continue to refine the type of $a$ after the aliasing,
as it still holds an \emph{open} table.
At the end of this example, $a$ has type
$\{``x":\String, ``y":\String \cup \Nil, ``z":\Integer\}_{open}$.

>>>>

Multiple assignments also appear in function applications, and the
rule \textsc{T-APPLY1} uses the same behavior from multiple assignments
to drop extra parameters and replace missing parameters:
\[
\begin{array}{c}
\mylabel{T-APPLY1}\\
\dfrac{\env_{1} \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
       \env_{2} \vdash el:r_{3}, \env_{3} \;\;\;
       r_{3} \lesssim r_{1}}
      {\env_{1} \vdash e(el)_{m}:r_{2}, \env_{3}}
\end{array}
\]

Assuming that $x$ and $f$ are local variables that are in the environment
with types $\Integer \cup \Nil$ and
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*} \rightarrow \Integer{*}$,
respectively.
The program
\[
x = f(``foo")_{m}
\]
type checks through the rule \textsc{T-APPLY1} because
\[
``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]
and the program
\[
x = f(``foo",1,2,3)_{m}
\]
also type checks through the rule \textsc{T-APPLY1} because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]

However, we also expect from a type system that it catches arity mismatch.
For this reason, our type system also has different typing rules for
expression lists that appear in function applications.
These rules do not include $\Nil{*}$ when the tuple type does not end
in a variadic type.
The rule \textsc{T-APPLY1-STRICT} use these rules to type check
expression lists, and it also uses the predicate $adjust$ to replace
any missing value with $\Nil$:
\[
\begin{array}{c}
\mylabel{T-APPLY1}\\
\dfrac{\env_{1} \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
       \env_{2} \vdash el:r_{3}, \env_{3} \;\;\;
       r_{4} = adjust(r_{3}, r_{1}) \;\;\;
       r_{4} \lesssim r_{1}}
      {\env_{1} \vdash e(el)_{m}:r_{2}, \env_{3}}
\end{array}
\]

@@

Now, assuming that $f$ has type
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void \rightarrow \Integer{*}$,
the program
\[
x = f(``foo")_{m}
\]
type checks through the rule \textsc{T-APPLY1-STRICT} because
\[
``foo" \times \Nil \times \Nil \times \Void \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]
but the program
\[
x = f(``foo",1,2,3)_{m}
\]
does not type check through the rule \textsc{T-APPLY1-STRICT} because
\[
``foo" \times 1 \times 2 \times 3 \times \Void \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]

The same happens with the rules that type check the \textbf{return}
statement.
The rule \textsc{T-RETURN} uses $\Nil{*}$ to replace eventual missing
values, while the rule \textsc{T-RETURN-STRICT} uses the predicate
\emph{adjust} to replace values that are certainly missing:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-RETURN}\\
\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2}(\ret) = r_{2} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1} \vdash \mathbf{return} \; el:\env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-RETURN-STRICT}\\
\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2}(\ret) = r_{2} \;\;\;
       r_{3} = adjust(r_{1}, r_{2}) \;\;\;
       r_{3} \lesssim r_{2}}
      {\env_{1} \vdash \mathbf{return} \; el:\env_{2}}
\end{array}
\end{array}
\]

Now, we move the discussion to the rules that type check the
different table constructors that our abstract syntax includes,
as they will appear in the next examples:

The simplest form of the table constructor is the empty table $\{\}$.
Its rule is straightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env \vdash \{\}:\{\}_{unique}, \env
\end{array}
\]

Our abstract syntax reduces the more complex uses of the table
constructor into three forms: $\{\;\vec{p}\;\}$, $\{\;{...}_{m}\;\}$,
and $\{\;\vec{p},{...}_{m}\;\}$.
The first one uses a list of table fields $[e_{1}] = e_{2}$,
the second one uses just the vararg expression, and the third one uses both.
We did not include in the abstract syntax a table constructor
that uses a list of expressions because we can write it using the first form.
For instance, we can write the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.
We chose to not include the latter form because its type checking rule is similar
to the type checking rule of the first form, and just controls the type of
the omitted indices.
For the same reason we did not include table constructors with
multiple expressions, which can return a tuple type that does not end
with a variadic type.
We did include the vararg expression in the last position to show how
our type system can type check a table type that has a record part and
also an array part.

The rule \textsc{T-CONSTRUCTOR2} type checks a table type of the first form:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       t = wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique})}
      {\env_{1} \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
\end{array}
\]

The rule \textsc{T-CONSTRUCTOR2} type checks each table field $p_{i}$,
uses the type of the table fields to build the table type, and
uses the predicate \emph{wf} to check whether the table type is
well formed.
Formally, a table type is well-formed if it obeys the following rule:
\[
\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
\]

The type checking of the table fields would be straightforward
if our table types allowed any first-level type in the keys.
The rules \textsc{T-FIELD1}, \textsc{T-FIELD2}, and \textsc{T-FIELD3}
check if the type of the key is a subtype of $\Boolean$, $\Number$,
or $\String$, respectively:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-FIELD1}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Boolean}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD2}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Number}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD3}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \String}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
\end{array}
\end{array}
\]

The rule \textsc{T-FIELD4} is a fallback for the previous rules:
\[
\begin{array}{c}
\mylabel{T-FIELD4}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3}}
      {\env_{1} \vdash [e_{1}] = e_{2}: \Value{:}close(t_{2}), \env_{3}}
\end{array}
\]

The rule \textsc{T-CONSTRUCTOR2} along with the rules for table fields
assign the type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ to
the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$.

Typed Lua handles arrays as hashes that map integers to some type $t$.
In Lua, programmers often use the vararg expression to initialize arrays.
The rules \textsc{T-CONSTRUCTOR3} and \textsc{T-CONSTRUCTOR4} define
the behavior of the vararg expression for this case:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-CONSTRUCTOR3}\\
\dfrac{\env_{1}({...}) = t}
      {\env_{1} \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env_{1}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-CONSTRUCTOR4}\\
\dfrac{\begin{array}{c}
       \env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       \env_{i+1}({...}) = t_{v}\\
       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t_{v} \cup \Nil\}_{unique}) = t_{t}
       \end{array}}
      {\env_{1} \vdash \{\;\vec{p},\;{...}_{m}\;\}:t_{t}, \env_{n+1}}
\end{array}
\end{array}
\]

The rule \textsc{T-CONSTRUCTOR3} type checks the case where we use
only the vararg expression inside a table constructor to initialize
an array.
If we assume that the $...$ is in the environment and has type $\String$,
the rule \textsc{T-CONSTRUCTOR3} gives the type $\{\Integer:\String \cup \Nil\}_{unique}$
to the table constructor $\{{...}_{m}\}$.

The rule \textsc{T-CONSTRUCTOR4} type checks the case where we use
the table constructor as a record that includes an array part.
If we assume that the $...$ is in the environment and has type $\String$,
the rule \textsc{T-CONSTRUCTOR4} gives the type
$\{``x":``foo", ``y":\mathbf{true}, \Integer:\String \cup \Nil\}_{unique}$
to the table constructor $\{[``x"] = ``foo", [``y"] = \mathbf{true}, {...}_{m}\}$.
As another example, the rule \textsc{T-CONSTRUCTOR4} fails to type check
the table constructor $\{ [1] = ``foo", [10] = \mathbf{true}, {...}_{m} \}$,
because the keys $1$ and $10$ overlap the array part.

After discussing the typing rules of the table constructor,
we start the discussion of the rules that define the most
unusual feature of our type system: the refinement of table types.
The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
and \emph{unique} table types:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\begin{array}{c}
       \env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique}\\
       \env_{1} \vdash c:k_{n+1}, \env_{2} \;\;\;
       \not \exists i \in 1..n \; t_{1} \lesssim k_{i} \;\;\;
       v_{n+1} = close(t)
       \end{array}}
      {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, k_{n+1}{:}v_{n+1}\}_{open|unique}]}
\end{array}
\]

@@

This example type checks because all the environment changes are consistent, and
$``foo" \times 1 \times \Nil{*} \lesssim \String \times \Integer \times \Value{*}$.

@@

Besides allowing the refinement of table types through field assignment,
Typed Lua also keeps track whether it is safe or not safe to change the
type of a table.
For instance,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{``foo":\Integer\}_{unique} = \{ [``foo"] = 5 \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``foo":\Integer \cup \Nil \}_{closed} = a \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$b[``foo"] = \mathbf{nil}$}
\end{tabular}
\end{center}
does not type check because aliasing $a$ produces the type
$\{``foo":\Integer\}_{closed}$ that is not subtype of
$\{``foo":\Integer \cup \Nil\}_{closed}$, the type of $b$.
The rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this behavior:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}}
      {\env_{1} \vdash n_{e}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}, \env_{1}[n \mapsto \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}]}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}}
      {\env_{1} \vdash n_{e}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}, \env_{1}}
\end{array}
\end{array}
\]

Aliasing \emph{unique} and \emph{open} table types can be unsafe.
For this reason, the rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2}
promote the type of an alias expression from \emph{unique} or \emph{open}
to \emph{closed}.
The rule \textsc{T-IDREAD1} also promotes the aliased variable from
\emph{unique} to \emph{open}, as \emph{unique} table types only
denote the type of table types that have no alias.

As another example,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = a \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
does not type check because we cannot add the field $``x"$ to $b$,
as its type is \emph{closed}, and thus does not allow changing the
value that is stored in the field $``x"$ of local $a$.

We also have different rules for type checking table indexing that
appear as expressions to avoid promoting an \emph{unique} table to
\emph{open} in these operations:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-INDEXREAD1}\\
\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\} \;\;\;
       \env_{1} \vdash e_{2}:t, \env_{2} \;\;\;
       \exists i \in 1{..}n \; t \lesssim k_{i}}
      {\env_{1} \vdash n_{e}[e_{2}]:v_{i}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-INDEXREAD1}\\
\dfrac{\env_{1} \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t, \env_{3} \;\;\;
       \exists i \in 1{..}n \; t \lesssim k_{i}}
      {\env_{1} \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
\end{array}
\end{array}
\]

Sometimes we want to use a local variable that holds an \emph{unique}
table type to initialize a \emph{closed} table type, as \emph{unique}
table types allow us to initialize optional fields.
To do that, Typed Lua includes a cast expression that allow us
to use an \emph{unique} table type before it becomes \emph{open}.
The rule \textsc{T-CAST} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-CAST}\\
\dfrac{\env_{1}(n) \subtype t}
      {\env_{1} \vdash {<}t{>} \; n:close(t), \env_{1}[n \mapsto t]}
\end{array}
\]

@@

We can use $a$ to initialize $b$ because the cast expression converts
the type of $a$ from $\{``x":\String, ``y":\String\}_{unique}$ to
$\{``x":\String, ``y":\String \cup \Nil\}_{open}$, and results in
$\{``x":\String, ``y":\String \cup \Nil\}_{closed}$.
We can continue to refine the type of $a$ after the aliasing,
as it still holds an \emph{open} table.
At the end of this example, $a$ has type
$\{``x":\String, ``y":\String \cup \Nil, ``z":\Integer\}_{open}$.

We also need to make sure to close all \emph{unique} and \emph{open}
table types before we type check another scope.
The rule \textsc{T-FUNCTION3} shows this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION3}\\
\dfrac{closeall(\env_{1}[\vec{n} \mapsto \vec{t}, \ret \mapsto r]) \vdash s:\env_{2}}
      {\env_{1} \vdash \mathbf{fun} \; (\vec{n{:}t}){:}r \; s:\vec{t} \rightarrow r, closeset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))}
\end{array}
\]
